Hierarchical Dispatching and Overriding

### 2017-10-05 NewVersion/sections/Formalization.tex
Change formalization.
Done revising formalization.

\begin{figure*}[t]
\saveSpaceFig
\begin{displaymath}
\begin{array}{l}
\begin{array}{llrl}
\text{Interfaces}   & IL & \Coloneqq & \interface{I}{I}{M} \\
\text{Methods}      & M  & \Coloneqq & \method{I}{m}{I_x}{x}{J}{e}  \mid
									   \absmethod{I}{m}{I_x}{x}{J} \\
\text{Expressions}  & e  & \Coloneqq & x \mid
e.m(\overline{e}) \mid
\new{I} \mid
e.I?m(e) \; \mid \; \angl{I}e \\
\text{Context}      & \Gamma & \Coloneqq & \overline{x}:\overline{I} \\
\text{Values}       & v & \Coloneqq & \angl{I} \new{J} \\
%%\\
%%\text{Interface names} & I, J, K & & \\
%%\text{Method names} & m & & \\
%%\text{Variable names} & x & &
\end{array}
\end{array}
\end{displaymath}
\caption{Syntax of \name{}.}\label{fig:syntax}
\saveSpaceFig
\end{figure*}

@@

	\framebox{$ \judgeewf \Gamma {e:I} $} \hspace{.5in}
	\tvar \\
	\tinvk \\
	\tpathinvk \\
	% \tsuperinvk \\
	% \tstaticinvk  \\
	\tnew \\
	\tmethod \\
	\tabsmethod \\
	\tintf \\
	\tanno
\end{mathpar}
\saveSpaceFig
\caption{Typing and subtyping of \name{}.}
\label{fig:typingrules}
\end{figure*}

@@

\begin{figure*}[t]
\saveSpaceFig
\begin{mathpar}
	\sinvk \\
	\spathinvk \\
	% \ssuperinvk \\
	% \sstaticinvk \\
	\creceiver \hspace{.5in}
	\cpathreceiver \\
	%\cargs \\
	\cpathargs \\
	% \csuperargs \\
	% \cstaticargs \\
	\cstatictype \\
	\cfreduce \\
	\cannoreduce
\end{mathpar}
\caption{Small-step semantics.}\label{fig:smallstep}
\saveSpaceFig
\end{figure*}

@@

\begin{theorem}[Subject Reduction]~\label{theorem_subject}
$\textit{If } \ \judgeewf \Gamma {e : I} \; \textit{ and } \ e \rightarrow e',\ 
\textit{then } \judgeewf \Gamma {e' : I'} \ \textit{ for some } \ I' <: I.$
\end{theorem}
\begin{proof}
See Appendix~\ref{appendix_proof}.
\end{proof}

\begin{theorem}[Progress]~\label{theorem_progress}
$\textit{Suppose } e \textit{ is a well-typed expression, if } e \textit{ includes} $

\noindent $\left(\angl{J}\emph{\kwnew}\;I()\right).K?m(\overline{v}) \textit{ as a subexpression, then } \emph{\mbody}(m, I, K) = (I_0, \overline{I_x} \; \overline{x}, I_e\; e_0) $ $\textit{and } \num{\overline{x}} = \num{\overline{e}} \textit{ for some } \overline{I_x}, \overline{x}, I_e \textit{ and } e_0.$
\end{theorem}
\begin{proof}
See Appendix~\ref{appendix_proof}.
\end{proof}

\begin{theorem}[Type Soundness]~\label{theorem_soundness}
If $\judgeewf \o {e : I}$ and $e \to^* e'$ with $e'$ a normal form, then $e'$ is 
a value $v$ with $\judgeewf \o {v:J}$ and $J \subtype I$.
\end{theorem}
\begin{proof}
Immediate from Theorem~\ref{theorem_subject} and Theroem~\ref{theorem_progress}.
\end{proof}

>>>>

\begin{figure*}[t]
\saveSpaceFig
\begin{displaymath}
\begin{array}{l}
\begin{array}{llrl}
\text{Interfaces}   & IL & \Coloneqq & \interface{I}{I}{M} \\
\text{Methods}      & M  & \Coloneqq & \method{I}{m}{I_x}{x}{J}{e}  \mid
									   \absmethod{I}{m}{I_x}{x}{J} \\
\text{Expressions}  & e  & \Coloneqq & x \mid
e.m(\overline{e}) \mid
\new{I} \mid \; (I)e \\
\text{Context}      & \Gamma & \Coloneqq & \overline{x}:\overline{I} \\
\text{Values}       & v & \Coloneqq & (I) \new{J} \\
%%\\
%%\text{Interface names} & I, J, K & & \\
%%\text{Method names} & m & & \\
%%\text{Variable names} & x & &
\end{array}
\end{array}
\end{displaymath}
\caption{Syntax of \name{}.}\label{fig:syntax}
\saveSpaceFig
\end{figure*}

@@

	\framebox{$ \judgeewf \Gamma {e:I} $} \hspace{.5in}
	\tvar \\
	\tinvk \\
	% \tpathinvk \\
	% \tsuperinvk \\
	% \tstaticinvk  \\
	\tnew \\
	\tmethod \\
	\tabsmethod \\
	\tintf \\
	\tanno
\end{mathpar}
\saveSpaceFig
\caption{Typing and subtyping of \name{}.}
\label{fig:typingrules}
\end{figure*}

@@

\begin{figure*}[t]
\saveSpaceFig
\begin{mathpar}
	% \sinvk \\
	\spathinvk \\
	% \ssuperinvk \\
	% \sstaticinvk \\
	\creceiver \hspace{.5in}
	% \cpathreceiver \\
	\cargs \\
	% \cpathargs \\
	% \csuperargs \\
	% \cstaticargs \\
	\cstatictype \\
	\cfreduce \\
	\cannoreduce
\end{mathpar}
\caption{Small-step semantics.}\label{fig:smallstep}
\saveSpaceFig
\end{figure*}

@@

\begin{theorem}[Subject Reduction]~\label{theorem_subject}
If $\judgeewf \Gamma {e : I}$ and $e \rightarrow e'$, 
then $\judgeewf \Gamma {e' : I}$.
\end{theorem}
\begin{proof}
See Appendix~\ref{appendix_proof}.
\end{proof}

\begin{theorem}[Progress]~\label{theorem_progress}
Suppose $e$ is a well-typed expression, if $e$ includes 
$\left((J)\emph{\kwnew}\;I()\right).m(\overline{v})$ as a sub-expression, then $\mbody(m, I, J) = (I_0, \overline{I_x} \; \overline{x}, I_e\; e_0)$ and $\num{\overline{x}} = \num{\overline{v}}$ for some $I_0$, $\overline{I_x}$, $\overline{x}$, $I_e$ and $e_0$.
\end{theorem}
\begin{proof}
See Appendix~\ref{appendix_proof}.
\end{proof}

\begin{theorem}[Type Soundness]~\label{theorem_soundness}
If $\judgeewf \o {e : I}$ and $e \to^* e'$ with $e'$ a normal form, then $e'$ is 
a value $v$ with $\judgeewf \o {v:I}$.
\end{theorem}
\begin{proof}
Immediate from Theorem~\ref{theorem_subject} and Theroem~\ref{theorem_progress}.
\end{proof}

### 2017-10-05 NewVersion/sections/TypeSafety.tex
Done revising formalization.

\begin{lemma}[Method Type Preservation]~\label{lemma:mbody_type_preservation}
If $\mbody(m, I_0, J_0) = (K, \overline{I_x}\ \_, I_e\ \_)$, then for any $I'_0 \subtype I_0$, $\mbody(m, I'_0, J_0) = (K', \overline{I_x}\ \_, I_e\ \_)$.
\end{lemma}
\begin{proof}

Since $\mbody(m, I_0, J_0)$ is defined, by \textsc{(T-Intf)} we derive that $\mbody(m, I'_0, J_0)$ is also defined. Suppose that $$\mostSpecific(m, I_0, J_0) = \{I\}$$ $$\mostSpecificOverride(m, I_0, I) = \{K\}$$
$$\mostSpecific(m, I'_0, J_0) = \{I'\}$$ $$\mostSpecificOverride(m, I'_0, I') = \{K'\}$$

Below we use $I[m\uparrow J]$ to denote the type of method $m$ defined in $I$ that overrides $J$. We have to prove that $K'[m\uparrow I'] = K[m\uparrow I]$.
Two facts:
\begin{itemize}
	\item A. By $\textsc{(T-Intf)}$, $\canOverride$ ensures that an override between any two original methods preserves the method type. Formally, $$I_1 \subtype I_2 \ \ \Rightarrow\ \ I_1[m\uparrow I_1] = I_2[m\uparrow I_2]$$
	\item B. By $\textsc{(T-Method)}$ and $\textsc{(T-AbsMethod)}$, any partial override also preserves method type. Formally,
	  $$I_1 \subtype I_2\ \ \Rightarrow\ \ I_1[m\uparrow I_2] = I_2[m\uparrow I_2]$$
\end{itemize}

Since $K \subtype I, K' \subtype I'$, by Fact B, $$K[m\uparrow I] = I[m\uparrow I] \quad K'[m\uparrow I'] = I'[m\uparrow I']$$

Hence it suffices to prove that $I'[m\uparrow I'] = I[m\uparrow I]$. Since $I'_0 \subtype I_0$, by the definition of $\mostSpecific$ and $\prune$, $I'$ and $I$
appear in the same set that is pruned, hence $I' \subtype I$. By Fact A, the proof is done.
\end{proof}

@@

\textbf{Case PathInvk.} $e = e_0.J_0?m(\overline{e})$. By \textsc{(T-PathInvk)}, suppose
	$$\judgeewf {\Gamma, \overline{x}:\overline{I_x}} {e_0 : I_0} \quad I_0 \subtype J_0$$
	$$\mbody(m, I_0, J_0) = (\_, \overline{J}\ \_, I\ \_)$$
	$$\judgeewf {\Gamma, \overline{x}:\overline{I_x}} {\overline{e} : \overline{I_e}} \quad 
	  \overline{I_e} \subtype \overline{J} \quad \judgeewf {\Gamma, \overline{x}:\overline{I_x}} {e : I}$$

@@

Again by \textsc{(T-PathInvk)}, $\judgeewf {\Gamma} {[\overline{y}/\overline{x}]e : I}$.

@@

\subsubsection{Proof for Theorem~\ref{theorem_subject}}
\begin{proof} ~\\
\indent \textbf{Case S-Invk.} Let
   $$e = (\angl{J}\new{I}).m(\overline{v}) \quad \judgeewf {\Gamma} {e : I_e}$$
   $$e' = (\angl{J}\new{I}).J?m(\overline{v})$$

@@

At this moment, \textsc{(T-PathInvk)} is immediately applicable by the above conditions, including that $e'$ also invokes $\mbody(m, J, J)$. Hence $\judgeewf {\Gamma} {e' : I_e}$.

\textbf{Case S-PathInvk.} Let
  $$e = (\angl{J}\new{I}).K?m(\overline{v}) \quad \judgeewf {\Gamma} {e : I_e}$$
  $$e' = \angl{I_{e_0}}[\overline{\angl{I_x}} \overline{v}/\overline{x}, \angl{I_0}\new{I}/\kwthis]e_0$$
  $$\mbody(m, I, K) = (I_0, \overline{I_x} \; \overline{x}, I_{e_0} \; e_0)$$

Since $\mbody(m, I, K)$ is defined, the definition of $\mbody$ ensures that $I \subtype I_0$. And since $e$ is well-typed, by \textsc{(T-PathInvk)},
  $$\judgeewf {\Gamma} {\overline{v} : \overline{I_v}} \quad \overline{I_v} \subtype \overline{I_x}$$

By the rules \textsc{(T-Anno)} and  \textsc{(T-New)},
  $$\judgeewf {\Gamma} {\overline{\angl{I_x}}\overline{v} : \overline{I_x}} \quad \judgeewf {\Gamma} {\angl{I_0}\new{I} : I_0}$$

@@

Hence by Lemma~\ref{lemma:subst_type_preservation}, the substitution preserves typing, thus
  $$\judgeewf {\Gamma} {[\overline{\angl{I_x}} \overline{v}/\overline{x}, \angl{I_0}\new{I}/\kwthis]e_0 : I'_{e_0}}$$

Since $I'_{e_0} \subtype I_{e_0}$, the conditions of \textsc{(T-Anno)} are satisfied, hence $\judgeewf {\Gamma} {e' : I_{e_0}}$. Now we only need to prove that $I_{e_0} = I_e$. Since $I_{e_0}$ is from $\mbody(m, I, K)$, whereas $I_e$ is from $\mbody(m, J, K)$, by the rule  \textsc{(T-PathInvk)}. By $\textsc{(T-Anno)}$ and the condition that $e$ type-checks, we know that $I \subtype J$. By Lemma~\ref{lemma:mbody_type_preservation}, $I_{e_0} = I_e$.

@@

\textbf{Case C-PathReceiver.} Straightforward induction.

@@

\textbf{Case C-PathArgs.} Straightforward induction.

@@

Since $e$ is well-typed, by \textsc{(T-PathInvk)} and \textsc{(T-Anno)} we know that
$$I \subtype J\textrm{, and }\mbody(m, J, K) \textrm{ is defined}$$

By Lemma~\ref{lemma:mbody_type_preservation}, $\mbody(m, I, K)$ is also defined, and the type checker ensures the expected number of arguments.

On the other hand, $I \subtype K$. By the definition of $\mostSpecific$, $$\mostSpecific(m, I, K) \subseteq \mostSpecific(m, I, I)$$

By \textsc{(T-New)}, $\canOverride(I)$ should be true. By the definition of $\canOverride$, any $J\in\mostSpecific(m, I, I)$ satisfies that
$\mostSpecificOverride(m, I, J)$ contains only one interface, in which the $m$ that overrides $J$ is a concrete method. Therefore $\mbody(m, I, K)$ also provides a concrete method, which finishes the proof.
\end{proof}

>>>>

\begin{lemma}[Method Type Preservation]~\label{lemma:mbody_type_preservation}
If $\mbody(m, J, J) = (K, \overline{I_x}\ \_, I_e\ \_)$, then for any $I \subtype J$, $\mbody(m, I, J) = (K', \overline{I_x}\ \_, I_e\ \_)$.
\end{lemma}
\begin{proof}

%\begin{comment}	
Since $\mbody(m, J, J)$ is defined, by \textsc{(T-Intf)} we derive that $\mbody(m, I, J)$ is also defined. Suppose that $$\mostSpecific(m, J, J) = \{I_0\}$$ $$\mostSpecificOverride(m, J, I_0) = \{K\}$$
$$\mostSpecific(m, I, J) = \{I'_0\}$$ $$\mostSpecificOverride(m, I, I'_0) = \{K'\}$$

Below we use $I[m\uparrow J]$ to denote the type of method $m$ defined in $I$ that overrides $J$. We have to prove that $K'[m\uparrow I'_0] = K[m\uparrow I_0]$.
Two facts:
\begin{itemize}
	\item A. By $\textsc{(T-Intf)}$, $\canOverride$ ensures that an override between any two original methods preserves the method type. Formally, $$I_1 \subtype I_2 \ \ \Rightarrow\ \ I_1[m\uparrow I_1] = I_2[m\uparrow I_2]$$
	\item B. By $\textsc{(T-Method)}$ and $\textsc{(T-AbsMethod)}$, any partial override also preserves method type. Formally,
	  $$I_1 \subtype I_2\ \ \Rightarrow\ \ I_1[m\uparrow I_2] = I_2[m\uparrow I_2]$$
\end{itemize}

By definition of $\mostSpecificOverride$, $K \subtype I_0, K' \subtype I'_0$. By Fact B, $$K[m\uparrow I_0] = I_0[m\uparrow I_0] \quad K'[m\uparrow I'_0] = I'_0[m\uparrow I'_0]$$

Hence it suffices to prove that $I'_0[m\uparrow I'_0] = I_0[m\uparrow I_0]$. Actually when calculating $\mostSpecific(m, J, J)$, by the definition of $\mostSpecific$ we know that $I_0 \subtype J$ and $I_0[m \; \kwoverride \; I_0]$ is defined. So when calculating $\mostSpecific(m, I, J)$ with $I \subtype J$, $I_0$ should also appear in the set before pruned, since the conditions are again satisfied. But after pruning, only $I'_0$ is obtained, by definition of $\prune$ it implies $I'_0 \subtype I_0$. By Fact A, the proof is done.
%\end{comment}

@@

\begin{comment}
\textbf{Case PathInvk.} $e = e_0.J_0?m(\overline{e})$. By \textsc{(T-PathInvk)}, suppose
	$$\judgeewf {\Gamma, \overline{x}:\overline{I_x}} {e_0 : I_0} \quad I_0 \subtype J_0$$
	$$\mbody(m, I_0, J_0) = (\_, \overline{J}\ \_, I\ \_)$$
	$$\judgeewf {\Gamma, \overline{x}:\overline{I_x}} {\overline{e} : \overline{I_e}} \quad 
	  \overline{I_e} \subtype \overline{J} \quad \judgeewf {\Gamma, \overline{x}:\overline{I_x}} {e : I}$$

@@

Again by \textsc{(T-PathInvk)}, $\judgeewf {\Gamma} {[\overline{y}/\overline{x}]e : I}$.
\end{comment}

@@

\subsubsection{Proof for Theorem~\ref{theorem_subject}}
\begin{proof} ~\\
\indent \textbf{Case S-Invk.} Let
\begin{comment}
   $$e = (\angl{J}\new{I}).m(\overline{v}) \quad \judgeewf {\Gamma} {e : I_e}$$
   $$e' = (\angl{J}\new{I}).J?m(\overline{v})$$

@@

At this moment, \textsc{(T-PathInvk)} is immediately applicable by the above conditions, including that $e'$ also invokes $\mbody(m, J, J)$. Hence $\judgeewf {\Gamma} {e' : I_e}$.
\textbf{Case S-PathInvk.} Let
\end{comment}
  $$e = ((J)\new{I}).m(\overline{v}) \quad \judgeewf {\Gamma} {e : I_e}$$
  $$e' = (I_{e_0})[\overline{(I_x)v}/\overline{x}, (I_0)\new{I}/\kwthis]e_0$$
  $$\mbody(m, I, J) = (I_0, \overline{I_x} \; \overline{x}, I_{e_0} \; e_0)$$

Since $\mbody(m, I, J)$ is defined, the definition of $\mbody$ ensures that $I \subtype J$. And since $e$ is well-typed, by \textsc{(T-Invk)},
  $$\judgeewf {\Gamma} {\overline{v} : \overline{I_v}} \quad \overline{I_v} \subtype \overline{I_x}$$

By the rules \textsc{(T-Anno)} and  \textsc{(T-New)},
  $$\judgeewf {\Gamma} {\overline{(I_x)v} : \overline{I_x}} \quad \judgeewf {\Gamma} {(I_0)\new{I} : I_0}$$

@@

Hence by Lemma~\ref{lemma:subst_type_preservation}, the substitution preserves typing, thus
  $$\judgeewf {\Gamma} {[\overline{(I_x)v}/\overline{x}, (I_0)\new{I}/\kwthis]e_0 : I'_{e_0}}$$

Since $I'_{e_0} \subtype I_{e_0}$, the conditions of \textsc{(T-Anno)} are satisfied, hence $\judgeewf {\Gamma} {e' : I_{e_0}}$. Now we only need to prove that $I_{e_0} = I_e$. Since $I_{e_0}$ is from $\mbody(m, I, J)$, whereas $I_e$ is from $\mbody(m, J, J)$, by the rule \textsc{(T-Invk)} on $e$. Since $I \subtype J$, by Lemma~\ref{lemma:mbody_type_preservation}, $I_{e_0} = I_e$.

@@

%\textbf{Case C-PathReceiver.} Straightforward induction.

@@

%\textbf{Case C-PathArgs.} Straightforward induction.

@@

Since $e$ is well-typed, by \textsc{(T-Invk)} and \textsc{(T-Anno)} we know that
$$I \subtype J\textrm{, and }\mbody(m, J, J) \textrm{ is defined}$$

By \textsc{(T-Intf)}, $\mbody(m, I, J)$ is also defined, and the type checker ensures the expected number of arguments.

On the other hand, since $I \subtype J$, by the definition of $\mostSpecific$, $$\mostSpecific(m, I, J) \subseteq \mostSpecific(m, I, I)$$

By \textsc{(T-New)}, $\canOverride(I) = True$. By the definition of $\canOverride$, any $J_0\in\mostSpecific(m, I, I)$ satisfies that
$\mostSpecificOverride(m, I, J_0)$ contains only one interface, in which the $m$ that overrides $J_0$ is a concrete method. Therefore $\mbody(m, I, J)$ also provides a concrete method, which finishes the proof.
\end{proof}
